---
title: "Project 2"
author: "brian higginbotham"
date: "2023-10-06"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE)
```

## Summary

This project encompasses querying data from an API and manipulating the results to produce some insightful EDA's.

I chose the basketball API [Free NBA](https://rapidapi.com/theapiguy/api/free-nba) from [rapidapi](https://rapidapi.com) because I thought it would be interesting to compare player and team information in a way that gave the user more control than traditional basketball reference websites. As we will see, this did provide some interesting and frustrating results.  

## Required Packages

Here is a list of the packages that were used in this project.

```{r packages}
library(httr)
library(tidyverse)
library(dplyr)
library(tidyr)
library(jsonlite)
library(ggplot2)
library(gganimate)
library(gifski)
library(png)
```

## Query API

My plan for the API query was to create a call that would allow the user to pull player and game data. For the player data, I would allow the user to select a specific player and season to collect data on or they could collect data on a player's entire career. For the game data, I wanted to allow the user to select the team, season, and specify whether they wanted regular or postseason data.  

The endpoints of the API:  
  * `Players` - contains player name, id, physical info (height, weight, etc), drafted team name
  * `Teams` - contains team name, id, city, team abbreviation, team full name
  * `Games` - date of game, game id, team ids (home & visitor), scores (home & visitor), season, postseason
  * `Stats` - essentially the box score for each player on every game they've played (points, rebounds, etc)

There were a few obstacles to creating the API query that should be mentioned here since a large part of constructing the functions was to get around them.  

The first issue is that the Free NBA API appears to be designed for websites or apps and NOT for data mining. Results from the API query come back "paged" with only the first page being pulled. This makes sense if one is pulling the data in through a website or app and can click on "page 2" to pull the next page of data. But for my purposes, a loop needed to be written to collect each page of data at a time.  

The second issue was that the parameter options provided by the API were not very helpful. Specific players or teams could be searched for if one had the `player_id` or `team_id` - both of which were numeric keys that had no immediate reference.  

The first step to creating the `player_stats()` query was to write a small function that could take in the player's name and return the `player_id` to be used in the `Stats` endpoint.

```{r player_id}
player_id <- function(last_name, first_name = ""){
  
  url <- "https://free-nba.p.rapidapi.com/players"
  
  queryString <- list(
    page = "0",
    per_page = "25",
    search = paste(first_name, last_name)
  )
  
  response <- VERB("GET", url, query = queryString, add_headers('X-RapidAPI-Key' = '3009c8c91amsh65cad163db7085ap15d3d2jsnad2333f55e76', 'X-RapidAPI-Host' = 'free-nba.p.rapidapi.com'), content_type("application/octet-stream"))
  
  id <- fromJSON(rawToChar(response$content))$data[1]
  attributes(id)$names[1] <- "player_id"
  
  return(id)
}
```
<br>
Next, in order to loop through all the "pages" and collect the data, I'll need to know how many pages are associated with the queried player. Additionally, the API query will have the option of collecting data from one particular season or for a whole career, so the length of the data can vary considerably. Therefore, I set up a function to determine the total number of pages that would be returned based on the input parameters. This function is designed for a specific `year` parameter:
```{r page_num}
# The final function will have year default to "career" if it is left empty. For this piece, year will have a numeric value (2010, 2015, 2023, etc)
page_num <- function(last_name, first_name = "", year = "career"){
# This is the previous function that identifies the player_id
  id <- player_id(last_name, first_name)
  
  url <- paste0("https://free-nba.p.rapidapi.com/stats?seasons[]=", year, "&player_ids[]=", id)
  
  queryString <- list(
    page = "0",
    per_page = "25"
  )
  
  response <- VERB("GET", url, add_headers('X-RapidAPI-Key' = '3009c8c91amsh65cad163db7085ap15d3d2jsnad2333f55e76', 'X-RapidAPI-Host' = 'free-nba.p.rapidapi.com'), content_type("application/octet-stream"))
  
  n <- fromJSON(rawToChar(response$content))$meta$total_pages[[1]]
  return(n)
}
```
The input parameters for this function will be the same as the input parameters of the final function. In fact, this function returns the first "page" of our query, but since we don't know initially how many total pages the query is, we have to run this function first to get `n`, the total number of pages we need to loop through.  

If no year is specified in the query parameter, then the query will default to career stats. Since the URL in the above function is set up for "?seasons[]= year", I'll need to modify it slightly in order to get the total number of pages for the player's career stats.
```{r page_num_career}
page_num_career <- function(last_name, first_name = ""){
  id <- player_id(last_name, first_name)
# I've removed the season parameter from the URL
  url <- paste0("https://free-nba.p.rapidapi.com/stats?player_ids[]=", id)
  
  queryString <- list(
    page = "0",
    per_page = "25"
  )
  
  response <- VERB("GET", url, add_headers('X-RapidAPI-Key' = '3009c8c91amsh65cad163db7085ap15d3d2jsnad2333f55e76', 'X-RapidAPI-Host' = 'free-nba.p.rapidapi.com'), content_type("application/octet-stream"))
  
  n <- fromJSON(rawToChar(response$content))$meta$total_pages[[1]]
  return(n)
}
```
<br>
Now I'm ready to construct the query that will return the player's stats by year. As I mentioned above, it is very similar to the function that the returns the number of pages. The main difference is in that function I accessed the `$meta` component in the `response$content`, but now I will access the `$data` component.  

Let's take a look at the `player_stats_year()` function
```{r player_stats_year()}
player_stats_year <- function(last_name, first_name = "", year){
  id <- player_id(last_name, first_name)
  n <- page_num(last_name, first_name, year)
  stats <- data.frame()
  
  for(i in 1:n){
    url <- paste0("https://free-nba.p.rapidapi.com/stats?seasons[]=", year, "&player_ids[]=", id)
    
    queryString <- list(
      page = i,
      per_page = "25"
    )
    
    response <- VERB("GET", url, query = queryString, add_headers('X-RapidAPI-Key' = '3009c8c91amsh65cad163db7085ap15d3d2jsnad2333f55e76', 'X-RapidAPI-Host' = 'free-nba.p.rapidapi.com'), content_type("application/octet-stream"))
    
    temp <- fromJSON(rawToChar(response$content))$data
    attributes(temp$team)$names[1] <- "teamid"
    temp <- bind_cols(temp %>% select(-team), temp$team)
    attributes(temp$game)$names[1] <- "gameid"
    temp <- bind_cols(temp %>% select(-game), temp$game)
    attributes(temp$player)$names[1] <- "playerid"
    temp <- bind_cols(temp %>% select(-player), temp$player)
    
    temp <- temp %>% select("first_name", "last_name", "abbreviation", "date", "season", "home_team_id", "home_team_score", "visitor_team_id", "visitor_team_score", "ast", "blk", "min", "pf", "pts", "reb", "stl")
    stats <- rbind(stats, temp)
  }
  stats <- stats %>% mutate(min=substr(stats$min, 1, 2))
  stats$min <- as.numeric(stats$min)
  attributes(stats)$names[3] <- "team"
  
  return(stats)
}
```
After the `player_id()` and `page_num()` functions, I've initiated the empty dataframe `stats`. This is where I'll collect my pages of data from the `for loop`.  Next is the `for loop` in which I'll iterate the API call 1:n times, with each time being a successive page number. This is reflected in the `queryString` as page = i.  

After I collected the `response`, things got a little tricky. Initially, the `for loop` did not work because each successive page had the same index (1:25) and so it was throwing an error about rows already existing. So I decided the best way around this was to go ahead and select the relevant columns before it executed `rbind()`. However, this led to another issue - the returned `$data` data frame had subsetted data frames. Essentially, all the relevant Player, Team, and Game information from the API call were subsetted data frames. So in order to bring those subsetted data frames up to a column level that can be `selected()`, I used `bind_cols()`.  

Before executing `bind_cols()`, I had to rename the `id` column for each subsetted dataframe. Without this step, the function would return "rename notifications" regarding the `id` columns for each `bind_cols` on every page of the `for loop`.Essentially, the returned dataframe as well as the subsetted dataframes each have their own `id` column (`temp$id`, `game$id`, `team$id`, `player$id`). When `bind_cols()` is called, it pulls the subsetted dataframe columns up to the "primary" dataframe. In doing so, it drops the subsetted dataframe name and "attaches" the column name to the "primary" dataframe. So the `temp` dataframe then has multiple columns named `id`. R would automatically rename these columns (`id...1`, `id...18`, etc,) based on their position in the dataframe. In doing so, it would return a notification in the console.  For some function calls, it would return 180 -200 notifications.  

It would be worth noting that the functions did work even with the "rename notifications". But since they were part of the return, I thought it would be a good idea to figure out a way to prevent it. I'm not sure if this was the best way to go about it, but it did work.  

Finally, there were two different ways the `min` data was recorded - as an integer and in ms format (mm:ss). After some unsuccessful trials with `lubridate()` to get these two different formats uniform, I decided the seconds portion of the `min` variable was kind of unnecessary. So I removed the seconds portion and converted the entire column to numeric so it could be aggregated.  

I have a very similar function for the Player Stats by Career query. Exactly like the `page_num_career()` function, it simply has a modified URL.
```{r player_stats_career}
player_stats_career <- function(last_name, first_name = ""){
  id <- player_id(last_name, first_name)
  n <- page_num_career(last_name, first_name)
  stats <- data.frame()
  
  for(i in 1:n){
    url <- paste0("https://free-nba.p.rapidapi.com/stats?player_ids[]=", id)
    
    queryString <- list(
      page = i,
      per_page = "25"
    )
    
    response <- VERB("GET", url, query = queryString, add_headers('X-RapidAPI-Key' = '3009c8c91amsh65cad163db7085ap15d3d2jsnad2333f55e76', 'X-RapidAPI-Host' = 'free-nba.p.rapidapi.com'), content_type("application/octet-stream"))
    
    temp <- fromJSON(rawToChar(response$content))$data
    attributes(temp$team)$names[1] <- "teamid"
    temp <- bind_cols(temp %>% select(-team), temp$team)
    attributes(temp$game)$names[1] <- "gameid"
    temp <- bind_cols(temp %>% select(-game), temp$game)
    attributes(temp$player)$names[1] <- "playerid"
    temp <- bind_cols(temp %>% select(-player), temp$player)
    
    temp <- temp %>% select("first_name", "last_name", "abbreviation", "date", "season", "home_team_id", "home_team_score", "visitor_team_id", "visitor_team_score", "ast", "blk", "min", "pf", "pts", "reb", "stl")
    stats <- rbind(stats, temp)
  }
  stats <- stats %>% mutate(min=substr(stats$min, 1, 2))
  stats$min <- as.numeric(stats$min)
  attributes(stats)$names[3] <- "team"
  
  return(stats)
}
```
<br>
Finally, we can put it all together in one wrapper function that will determine which function to use: `player_stats_year()` or `player_stats_career()`
```{r stats_wrapper}
player_stats <- function(last_name, first_name = "", year = 'career'){
  if(year == "career"){
    return(player_stats_career(last_name, first_name))
  } else {
    return(player_stats_year(last_name, first_name, year))
  }
}
```
<br>
